---
title: "Assignment 3: Gut Microbiome Analysis (Vegan vs Omnivore)"
author: "Kexin Gong"
date: "2026-03-21"
---

# Introduction

The human gut microbiome plays a crucial role in metabolism, immune function, and overall health. Diet is one of the major factors influencing microbial composition. In this study, we compare gut microbiomes of vegan and omnivore individuals using shotgun metagenomic data.

# Methods

Raw sequencing data were obtained from the NCBI SRA and processed using SRA Toolkit. Taxonomic classification was performed using Kraken2 with the Standard-8 database. The resulting classification reports were analyzed in R.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
library(RColorBrewer)
```

# Sample groups

```{r}
omnivore_samples <- c("SRR8146935", "SRR8146936", "SRR8146938", "SRR8146956", "SRR8146957")

vegan_samples <- c("SRR8146944", "SRR8146963", "SRR8146968", "SRR8146951", "SRR8146952")

sample_info <- data.frame(
  sample = c(omnivore_samples, vegan_samples), 
  group = c(rep("Omnivore", length(omnivore_samples)),
            rep("Vegan", length(vegan_samples))
            )
  )

sample_info
```

# Load Kraken2 reports

```{r}
report_files <- list.files("../data", pattern = "\\.report$", full.names = TRUE)
report_files
```

# Read data

```{r}
read_kraken_report <- function(file) {
  df <- read.delim(file, header = FALSE)
  colnames(df) <- c("percent", "reads_clade", "reads_taxon", "rank_code", "taxid", "name")
  df$sample <- gsub("\\.report$", "", basename(file))
  df$name <- trimws(df$name)
  df
}

kraken_data <- lapply(report_files, read_kraken_report) %>%
  bind_rows()

head(kraken_data)
```

# Filter genus level

```{r}
genus_data <- kraken_data %>%
  filter(rank_code == "G") %>%
  select(sample, name, reads_clade)

head(genus_data)
```

# Create abundance table

```{r}
abundance_wide <- genus_data %>%
  pivot_wider(
    names_from = sample,
    values_from = reads_clade,
    values_fill = 0
  )

abundance_matrix <- abundance_wide %>%
  column_to_rownames("name") %>%
  as.matrix()

abundance_rel <- sweep(abundance_matrix, 2, colSums(abundance_matrix), "/")

abundance_rel[1:5, 1:5]
```

# Taxonomic composition

```{r fig.height=6, fig.width=11}
# Select the top 9 genera based on mean relative abundance across all samples
top_taxa <- rowMeans(abundance_rel) %>%
  sort(decreasing = TRUE) %>%
  head(10) %>%
  names()

# Convert the relative abundance matrix to long format
plot_data <- abundance_rel %>%
  as.data.frame() %>%
  rownames_to_column("taxon") %>%
  pivot_longer(-taxon, names_to = "sample", values_to = "abundance") %>%
  left_join(sample_info, by = "sample")

# Collapse low-abundance genera into an "Other" category
plot_data$taxon <- ifelse(plot_data$taxon %in% top_taxa,
                          plot_data$taxon,
                          "Other")

# Sum abundances after collapsing taxa
plot_data <- plot_data %>%
  group_by(sample, group, taxon) %>%
  summarise(abundance = sum(abundance), .groups = "drop")

# Preserve sample order
plot_data$sample <- factor(plot_data$sample, levels = sample_info$sample)

# Preserve taxon order
taxon_order <- c(top_taxa, "Other")
plot_data$taxon <- factor(plot_data$taxon, levels = taxon_order)

# Create stacked bar plot
cool_colors <- c("#81A6C6", "#BDA6CE", "#AACDDC", "#D2C4B4", "#F3E3D0", "#8E977D", "#9F8383", "#D2DCB6", "#758A93", "#B6AE9F", "#BFC6C4")

ggplot(plot_data, aes(x = sample, y = abundance, fill = taxon)) +
  geom_bar(stat = "identity", width = 0.85) +
  facet_grid(~ group, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = cool_colors) +
  labs(
    title = "Genus-level taxonomic composition",
    x = NULL,
    y = "Relative abundance",
    fill = "Genus"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```

# Alpha diversity

```{r}
alpha_df <- data.frame(
  sample = colnames(abundance_rel),
  shannon = diversity(t(abundance_rel), index = "shannon")
) %>%
  left_join(sample_info, by = "sample")

alpha_df
```

```{r fig.height=5, fig.width=6}
ggplot(alpha_df, aes(x = group, y = shannon, color = group)) +
  geom_boxplot(width = 0.3, outlier.shape = NA, alpha = 0.5, linewidth = 0.8) +
  geom_jitter(width = 0.08, size = 3, alpha = 1) +
  scale_color_manual(values = c("Omnivore" = "#E2A16F", 
                                "Vegan" = "#777C6D")) +
  theme_classic(base_size = 13) +
  labs(
    x = NULL,
    y = "Shannon diversity"
  ) +
  theme(
    legend.position = "none"
  )
```

# Statistics

```{r}
wilcox.test(shannon ~ group, data = alpha_df)
```

# Beta diversity

```{r}
bray_dist <- vegdist(t(abundance_rel), method = "bray")
pcoa <- cmdscale(bray_dist, k = 2)

pcoa_df <- data.frame(
  sample = rownames(pcoa),
  PC1 = pcoa[,1],
  PC2 = pcoa[,2]
) %>%
  left_join(sample_info, by = "sample")

pcoa_df
```

```{r fig.height=5, fig.width=6}
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Beta diversity (PCoA)")
```

# Differential abundance

```{r}
diff_results <- genus_data %>%
  left_join(sample_info, by = "sample") %>%
  group_by(name) %>%
  summarise(
    p_value = tryCatch(
      wilcox.test(reads_clade ~ group)$p.value,
      error = function(e) NA_real_
    ),
    vegan_mean = mean(reads_clade[group == "Vegan"], na.rm = TRUE),
    omnivore_mean = mean(reads_clade[group == "Omnivore"], na.rm = TRUE)
  ) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adj)

head(diff_results, 20)
```

# Results

Taxonomic composition analysis revealed variation in genus-level abundance across samples. Alpha diversity indicated differences in microbial diversity between vegan and omnivore groups. Beta diversity analysis suggested distinct clustering patterns. Differential abundance analysis identified genera with significant differences between dietary groups.

# Discussion

These results suggest that dietary patterns influence gut microbiome composition. Differences in microbial diversity and abundance may reflect dietary intake differences. However, limitations include small sample size and incomplete classification of reads.
